Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-11 demystify.tex
updates

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}

\section*{Who is this document for?}

@@

@@

\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\end{table}

>>>>

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
\section{Why the US and not other countries?}
TODO

\section{Who is this document for?}

@@

\section{What's the difference between PhD and MS in CS?}\label{sec:phd-vs-ms}

While both are graduate degrees, they are different in terms of objective, admission requirements, course requirements, duration, and funding as summarized in \autoref{tab:phd-vs-ms}.

\begin{table}
  \caption{PhD vs. MS}\label{tab:phd-vs-ms}
  \begin{tabular}{c|c|c}
  \toprule
  \textbf{} & \textbf{PhD} & \textbf{MS} \\
  \midrule
  Objective & Industry & Research \\
  Admission Req & Research experience & No research experience \\
  Coursework Req & Yes & Yes (research is much more important)  \\
  Duration & 2 years & 5--7 years \\
  Adviser Req & No & Yes \\
  Funding & No & Yes \\
  \bottomrule
  \end{tabular}
\end{table}

  \item Objective: an MS is typically to prepare you for industry, while a PhD is to prepare you for research. Some MS has thesis option but the research is much lighter compared to a PhD, which is research-focused.

  \item Admission requirements: MS also requires a good GPA, LoRs, SOP, and test scores, but research experience is not required.  PhD requires all of these, but research experience is a must. Moreover, MS programs are typically much less competitive than PhD programs. In fact, many MS program are viewed as a revenue source (\emph{cash cow}) for the department, as they are not funded and students pay tuition.

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation (which is typically way harder and takes much longer time than coursework).

  \item Duration: an MS typically takes 2 years while a PhD takes 5--7 years (or even longer).  Many students get an MS along the way to a PhD, e.g., after finishing the 2-year course work.

  \item Adviser: MS students typically do not have an adviser (if you do thesis option then you will have one), while PhD students need an adviser who guides them in their research.

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (e.g., through TA or RA as explained in \autoref{sec:funding}).

\end{itemize}

@@

\begin{table}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}
